---
title: "Import ontology files"
author: "Zuguang Gu ( z.gu@dkfz.de )"
date: '`r Sys.Date()`'
output: html_document
vignette: >
  %\VignetteIndexEntry{2. Import ontology files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, echo = FALSE, message = FALSE}
library(knitr)
knitr::opts_chunk$set(
    error = FALSE,
    tidy  = FALSE,
    message = FALSE,
    warning = FALSE,
    fig.align = "center")
```


## The .obo format

There are several formats for ontology data. The most compact and readable
format is the `.obo` format, which was initially developed by GO. A lot of
ontologies in `.obo` format can be found from the [OBO
Foundry](http://obofoundry.org/) or
[BioPortal](https://bioportal.bioontology.org/). A description of the `.obo`
format can be found from
https://owlcollab.github.io/oboformat/doc/GO.format.obo-1_4.html.

In the **simone** package, the function `import_obo()` can be used to import
an `.obo` file to an `ontology_DAG` object. The input is a path on local
computer or an URL. In the following example, we use the [Plant
Ontology](http://obofoundry.org/ontology/po.html) as an example.


The link of `po.obo` can be found from that web package. You can download it or
directly provide it as an URL.

```{r, warning = FALSE}
library(simone)
dag1 = import_obo("https://raw.githubusercontent.com/Planteome/plant-ontology/master/po.obo")
dag1
```

There are also several meta columns attached to the object, such as
the name and the long definition of terms in the ontology.

```{r}
head(mcols(dag1))
```

Note rows in `mcols(dag1)` corresponds to terms in `dag_all_terms(dag)`.

The `is_a` relation between classes is of course saved in the DAG object
(specified in the `is_a` tag in the `.obo` file). Additional relation types
can also be selected (specified in the `relationship` tag). By default only the
relation type `part_of` is used. You can check other values associated with
the `relationship` tag and the `[Typedef]` section in the `.obo` file to select
proper additional relation types. Just make sure that the selected relation
types are transitive and not inversed (e.g. you cannot select `has_part` which is
a reversed relation of `part_of`).

Relations can also have a DAG structure. In `import_obo()`, if a parent
relation type is selected, all its offspring types are automatically selected.
For example, in GO, besides relations of `is_a` and `part_of`, there are also
`regulates`, `positively_regulates` and `negatively_regulates`, where the latter ones are child
relations of `regulates`. So if
`regulates` is selected as an additional relation type, the other two are
automatically selected.


```r
import_obo("file_for_go.obo", relation_type = c("part_of", "regulates"))
```

Finally, all the spaces specified in `relation_type` will be converted to
underlines. So it is the same if you specify `"part of"`.

## Other ontology formats

For ontologies in other formats, **simone** uses an external tool
[**ROBOT**](http://robot.obolibrary.org/) to convert them to `.obo` format and
internally uses `import_obo()` to import them. The file `robot.jar` is needed and
it can be downloaded from https://github.com/ontodev/robot/releases (Since
this is a tool in Java, you should have Java already available on the
machine).

The file `po.owl` can also be found from the [Plant
Ontology](http://obofoundry.org/ontology/po.html) web page.

```{r, eval = identical(Sys.info()["user"], "guz")}
dag2 = import_ontology("https://raw.githubusercontent.com/Planteome/plant-ontology/master/po.owl", 
    robot_jar = "~/Downloads/robot.jar")
```

```{r, eval = FALSE}
dag2
```

```{r, echo = FALSE}
if(identical(Sys.info()["user"], "guz")) {
    print(dag2)
} else {
    cat(
"An ontology_DAG object:
  Source: po, releases/2021-08-13
  1654 terms / 2510 relations
  Root: _all_
  Terms: PO:0000001, PO:0000002, PO:0000003, PO:0000004, ...
  Max depth: 13
  Aspect ratio: 24.85:1 (based on the longest distance to root)
                39.6:1 (based on the shortest distance to root)
  Relations: is_a, part_of

With the following columns in the metadata data frame:
  id, short_id, name, namespace, definition
")
}
```

The number of terms and relations are the same as in `dag1`.

More conveniently, the path of `robot.jar` can be set as a global option:

```r
simone_opt$robot_jar = "~/Downloads/robot.jar"
import_ontology("https://raw.githubusercontent.com/Planteome/plant-ontology/master/po.owl")
```

**ROBOT** supports the following ontology formats and they are automatically
identified according to the file contents.

- `json`: OBO Graphs JSON
- `obo`: OBO Format
- `ofn`: OWL Functional
- `omn`: Manchester
- `owl`: RDF/XML
- `owx`: OWL/XML
- `ttl`: Turtle 

## The .owl format

For some huge ontologies, it requires a huge amount of memory for **ROBOT** to
convert to the `.obo` format. If the ontology is in the `.owl` format (in the
RDF/XML seriation format), the function `import_owl()` can be optionally used.
`import_owl()` directly parses the `.owl` file and returns an `ontology_DAG` object.
The `import_owl()` is written from scratch and it is recommended to use only when `import_ontology()`
does not work.


```{r}
dag3 = import_owl("https://raw.githubusercontent.com/Planteome/plant-ontology/master/po.owl")
dag3
```

As you see, all the three DAG objects contain the same numbers of terms and relations.

## Session info

```{r}
sessionInfo()
```
