---
title: "Functional enrichment"
author: "Zuguang Gu ( z.gu@dkfz.de )"
date: '`r Sys.Date()`'
output: 
  html_vignette:
    css: main.css
    toc: true
vignette: >
  %\VignetteIndexEntry{10. Functional enrichment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE, message = FALSE}
library(knitr)
knitr::opts_chunk$set(
    error = FALSE,
    tidy  = FALSE,
    message = FALSE,
    warning = FALSE,
    fig.align = "center")
```

```{r, echo = FALSE}
knitr::knit_hooks$set(pngquant = knitr::hook_pngquant)

knitr::opts_chunk$set(
  dev = "ragg_png",
  fig.align = "center",
  pngquant = "--speed=10 --quality=50"
)
```

In many applications, semantic similarity analysis is integerated with gene
set enrichment analysis, especially taking GO as the source of gene sets.
**simona** provides functions that import ontologies already integrated with
gene annotations. **simona** also provides functions for over-representation analysis (ORA)
and functions to integrate the ORA results with semantic similarity analysis.

## GO


To add gene annotations for GO, just set the name of the "org.db" package for the specific organism.
For example "org.Hs.eg.db" for human and "org.Mm.eg.db" for mouse. The full list of supported
"org.db" packages can be found at https://bioconductor.org/packages/release/BiocViews.html#___AnnotationData (search `org.`).

```{r}
library(simona)
dag = create_ontology_DAG_from_GO_db(org_db = "org.Hs.eg.db")
dag
```

As the object `dag` prints, the genes stored in `dag` are in the EntreZ ID type. So when doing ORA,
the input gene list should also be in the EntreZ ID type.

We generate a list of random genes for testing:

```{r}
set.seed(123)
genes = random_items(dag, 500)
head(genes)
```

To perform ORA, use the function `dag_enrich_on_genes()`.

```{r}
tb = dag_enrich_on_genes(dag, genes)
tb = tb[order(tb$p_adjust), ]
head(tb)
```

We can take the significant GO terms and look at their semantic similarities. 

```{r}
top_go_ids = tb$term[1:100]
mat = term_sim(dag, top_go_ids)
```

```{r, fig.width = 6.5, fig.height = 6}
library(ComplexHeatmap)
Heatmap(mat, name = "similarity", 
    show_row_names = FALSE, show_column_names = FALSE,
    show_row_dend = FALSE, show_column_dend = FALSE)
```

And the significant GO terms on the global circular plot:

```{r, fig.width = 9, fig.height = 7}
dag_circular_viz(dag, top_go_ids)
```

One of the use of the semantic similarity matrix is to cluster GO terms in groups, to simplify the read of the results.
Here the semantic similarity matrix can be directly sent to `simplifyEnrichment()` function from the **simplifyEnrichment**
package. Since the terms are from GO, there will be word cloud associated with the heatmap to show their generl
biological functions in each cluster.

```{r, fig.width = 8.5, fig.height = 4.34}
library(simplifyEnrichment)
simplifyEnrichment(mat)
```

In the previous example, when setting the organism, we use the name of the
`org.db` package. The value can also directly be an `OrgDb` object. This expands
the use of the function since there are many `OrgDb` objects for less-studied organims available
on **AnnotationHub**.

The following code demonstrates the use of the delphin organism (Delphinus truncatus). `AH112417` is the ID of
this dataset. Please refer to **AnnotationHub** for the usage of the package.

```{r, eval = FALSE}
library(AnnotationHub)
ah = AnnotationHub()
org_db = ah[["AH112417"]]
dag = create_ontology_DAG_from_GO_db(org_db = org_db)
```

## Other ontologies

Besides GO, there are also other ontologies that have gene annotations integrated. 

### UniProt Keywords

UniProt Keywords (https://www.uniprot.org/keywords) is a set of controlled vocabulary developed in UniProt to describe
the biological functions of proteins. It is organised in a hierarchical way, thus in a form of the ontology. The function
`ontology_kw()` can import the UniProt Keywords ontology with gene annotations from a specific organims.

The function internally uses the **UniProtKeywords** package. All supported organisms can be found in the documentation
of `UniProtKeywords::load_keyword_genesets()`.

```{r}
dag = ontology_kw("human")
dag
```

As `dag` shows, the gene ID type is EntreZ ID. Similar as GO, we randomly generate a list of genes and perform ORA.

```{r}
genes = random_items(dag, 500)
tb = dag_enrich_on_genes(dag, genes)
tb = tb[order(tb$p_adjust), ]
top_go_ids = tb$term[1:50]
```

Obtain the semantic similarity matrix and make plots:

```{r, fig.width = 6.5, fig.height = 6}
mat = term_sim(dag, top_go_ids)
Heatmap(mat, name = "similarity", 
    show_row_names = FALSE, show_column_names = FALSE,
    show_row_dend = FALSE, show_column_dend = FALSE)
```

```{r, fig.width = 9, fig.height = 7}
dag_circular_viz(dag, top_go_ids)
```

We also also use `simplifyEnrichment()` to cluster terms in `mat`, but there is no word cloud around the heatmap.

```{r, fig.width = 5.5, fig.height = 4.34}
cl = simplifyEnrichment(mat)
head(cl)
```


### Ontologies from RGD

The following ontologies as well as the gene annotations are from the Rat Genome Database (RGD).
Although the RGD is a database for mouse, it also provides gene annotations for other oganisms.
The specific files used in each function can be found at https://download.rgd.mcw.edu/ontology/.

Note that the following functions may support different sets of organims. Please go to the documentations for the list.

**Pathway Ontology**

```{r}
dag = ontology_pw("human")
dag
```

Note that, in the pathway ontology, genes are saved in gene symbols.

**Chemical Entities of Biological Interest**

```{r, eval = FALSE}
dag = ontology_chebi("human")
```

**Disease Ontology**

```{r, eval = FALSE}
dag = ontology_rdo("human")
```

**Vertebrate Trait Ontology**

```{r, eval = FALSE}
dag = ontology_vt("human")
```

## Session info

```{r}
sessionInfo()
```
