---
title: "Term similarity"
author: "Zuguang Gu ( z.gu@dkfz.de )"
date: '`r Sys.Date()`'
output: html_document
vignette: >
  %\VignetteIndexEntry{5. Term similarity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE, message = FALSE}
library(knitr)
knitr::opts_chunk$set(
    error = FALSE,
    tidy  = FALSE,
    message = FALSE,
    warning = FALSE,
    fig.align = "center")
```

```{r, echo = FALSE}
knitr::knit_hooks$set(pngquant = knitr::hook_pngquant)

knitr::opts_chunk$set(
  dev = "ragg_png",
  fig.align = "center",
  pngquant = "--speed=10 --quality=30"
)
```

<style>
h1:not(.title) {
    color:grey;
    text-decoration: underline;
}
</style>

The methods of semantic similarity implemented in **simona** are mainly from
the [supplementary file](https://oup.silverchair-cdn.com/oup/backfile/Content_public/Journal/bib/18/5/10.1093_bib_bbw067/2/bbw067_supplementary_file.pdf?Expires=1692176766&Signature=Ks4C96x-kinySVP7rzVzMn8cGnQWXOqkB-6gYHhs48Bl9A2Vj~-mcI9pEnvntNBvjBD1hp4ZncMWUNT-y1ArrrQgDJ0dCCEJp5LG0Ag2oPWsC5NFEeUQiYNZCXal8Kwa3Ze1vtDHjIwZpT9-sZhn9ryqKHd3C88RvuGx10jcSYRL1Pec0F~a9CnrbivieCfFELg2OQTf5zpAobxhlqzWqv~sa1b62PunNZ~XkmnTGOJoSMQFnNaBy~bDNEf9Pd6eRYaOc8MUh5Br4xB4y8mivT6Et2vKzLrGJFQjYpvouQpBu0nf03xwHVmNFN5-6LsDnpyQPCy4nQ5DHYltVHkEOA__&Key-Pair-Id=APKAIE5G5CRDK6RD3PGA)
of the paper ["Mazandu et al., Gene Ontology semantic similarity tools: survey
on features and challenges for biological knowledge discovery. Briefings in
Bioinformatics 2017"](https://doi.org/10.1093/bib/bbw067). Original
denotations have been slightly modified to make them more consistent. Also
more explanations have been added in this vignette. Credits go to
the original paper.

## Denotations

The following denotations will be used throughout the vignette.

|  Denotation |  Description    |
| :----- | :------------------------ |
| $r$ | The root term of the DAG.  |
| $\delta(x)$ | The depth of a term $x$ in the DAG, which is the longest distance from root $r$. |
| $\delta_s(x)$ | The length of the longest path from root $r$ to a term $x$ via term $s$. |
| $\delta_\max$ | The maximal depth in the DAG. |
| $\eta(x)$ | The height of term $x$ in the DAG, which is the longest finite distance to leaf terms. |
| $\mathcal{C}_s$ | The set of child terms of term $s$. |
| $\mathcal{P}_s$ | The set of parents terms of term $s$. |
| $\mathcal{A}_s$ | The set of ancestor terms of term $s$. |
| $\mathcal{A}_s^+$ | The set of ancestor terms of term $s$, including $s$ itself. |
| $\mathcal{D}_s$ | The set of offspring terms of term $s$. |
| $\mathcal{D}_s^+$ | The set of offspring terms of term $s$, including $s$ itself. |
| $\mathcal{L}_s$ | The set of leaf terms that term $s$ can reach. |
| $\left| A \right|$ | Number of elements in set $A$. |
| $D_\mathrm{sp}(a, b)$ | The shortest distance bewteen $a$ and $b$. | 
| $\mathrm{len}(a, b)$ | The longest distance bewteen $a$ and $b$. |
| $\mathrm{len}_s(a, b)$ | The length of the longest path from $a$ and $b$ via $s$. |
| $\mathrm{CA}(a, b)$ | The set of common ancestors of term $a$ and $b$, i.e. $\mathrm{CA}(a, b) = \mathcal{A}_a^+ \cap \mathcal{A}_b^+$ |
| $\mathrm{LCA}(a, b)$ | Lowest common ancestor of $a$ and $b$, which is the common ancestor with the largest depth in DAG, i.e. $$\operatorname*{argmax}_{t \in \mathrm{CA}(a, b)} \delta(t)$$ There might be more than one LCA terms for given two terms, to simplify the calculation, the one with the smallest distance to $a$ and $b$ is used. |
| $\mathrm{NCA}(a, b)$ | Nearest common ancestor of $a$ and $b$, i.e. $$\operatorname*{argmin}_{t \in \mathrm{CA}(a, b)} \left( D_\mathrm{sp}(t, a) + D_\mathrm{sp}(t, b) \right)$$ If there are more than one NCA terms, the one with the highest depth is used. |
| $\mathrm{MICA}(a, b)$ | Most informative common ancestor of $a$ and $b$, i.e. $$\operatorname*{argmax}_{t \in \mathrm{CA}(a, b)} \left( \mathrm{IC}(t) \right )$$ There might be more than one MICA terms for given two terms, the one with the smallest distance to $a$ and $b$ is used.|
| $G_s$ | The set of annotated items on term $s$. |

Assume term $a$ is an ancestor of term $b$, $D_\mathrm{sp}(a, b)$ (the order
of $a$ and $b$ does not matter) is the normal shortest distance from $a$ to
$b$ in a directed graph. The definition is similar for $\mathrm{len}(a, b)$.

If term $a$ and $b$ are not in offspring/ancestor relationship, i.e. $a$ is
not an ancestor of $b$, and $b$ is not an ancestor of $a$, then

$$
\begin{align*}
D_\mathrm{sp}(a, b) &= \min_{t \in \mathrm{CA}(a, b)} \left( D_\mathrm{sp}(t, a) + D_\mathrm{sp}(t, b) \right) \\
\mathrm{len}(a, b) &= \max_{t \in \mathrm{CA}(a, b)} \left( \mathrm{len}(t, a) + \mathrm{len}(t, b) \right)
\end{align*}
$$


The wrapper function `term_sim()` calculates semantic similarities between terms in
the DAG with a specific method. Note the method name can be partially matched.

<pre class="r">
term_sim(dag, terms, method = ..., control = list(...))
</pre>

# IC-based or node-based methods

This type of methods considers a special ancestor term $c$ of terms $a$ and
$b$, which has the highest IC among all $a$ and $b$'s ancestor terms. Term $c$
is called the most informative common ancestor (MICA) which can be given by:

$$ \mathrm{IC}(c) = \max_{t \in \mathcal{A}_a^+ \cap \mathcal{A}_b^+} \mathrm{IC}(t)  $$

So if two terms are identical, MICA is the term itself, and if two terms have
ancestor/offspring relationship, MICA is the ancestor term.

In the following sections, if not specially mentioned, $c$ is always referred
to the MICA of $a$ and $b$.

## Sim_Lin_1998

The similarity is calculated as the IC of the MICA term $c$ normalized by the
average of the IC of the two terms:

$$ \mathrm{Sim}(a, b) = \frac{\mathrm{IC}(c)}{(\mathrm{IC}(a) + \mathrm{IC}(b))/2} = \frac{2 * \mathrm{IC}(c)}{\mathrm{IC}(a) + \mathrm{IC}(b)} $$


<pre class="r">
term_sim(dag, terms, method = "Sim_Lin_1998")
</pre>

Paper link: https://dl.acm.org/doi/10.5555/645527.657297.


## Sim_Resnik_1999

IC of the MICA term itself $\mathrm{IC}(c)$ can be a measure of how similar
two terms are, but its range is not in `[0, 1]`. There are several ways to
normalize $\mathrm{IC}(c)$ to the range of `[0, 1]`. Note some of the
normalization methods are restricted to _IC_annotation_ as the IC method.

### Nunif

It is normalized to the possible maximal IC value where a term only has one
item annotated.

$$ \mathrm{Sim}(a, b) = \frac{\mathrm{IC}(c)}{-\log(1/N)} = \frac{\mathrm{IC}(c)}{\log N} $$


where $N$ is the total number of items annotated to the whole DAG.

### Nmax

It is similar to _Nunif_, but normalized to the maximal IC of all terms in the
DAG. If there is a term with only one item annotated, _Nmax_ is identical to
the _Nunif_ method.

$$ \mathrm{Sim}(a, b) = \frac{\mathrm{IC}(c)}{\mathrm{IC}_\mathrm{max}} $$


### Nunivers

$\mathrm{IC}(c)$ is normalized by the maximal IC of term $a$ and $b$.

$$ \mathrm{Sim}(a, b) = \frac{\mathrm{IC}(c)}{\max \{ \mathrm{IC}(a), \mathrm{IC}(b) \}} $$

Paper link: https://doi.org/10.1613/jair.514, https://doi.org/10.1186/1471-2105-9-S5-S4, https://doi.org/10.1186/1471-2105-11-562, https://doi.org/10.1155/2013/292063.


The normalization method can be set with the `norm_method` parameter:

<pre class="r">
term_sim(dag, terms, method = "Sim_Resnik_1999", control = list(norm_method = "Nunif"))
term_sim(dag, terms, method = "Sim_Resnik_1999", control = list(norm_method = "Nmax"))
term_sim(dag, terms, method = "Sim_Resnik_1999", control = list(norm_method = "Nunivers"))
</pre>

## Sim_FaITH_2010

It is calculated as:

$$ \mathrm{Sim}(a, b) = \frac{\mathrm{IC}(c)}{\mathrm{IC}(a) + \mathrm{IC}(b) - \mathrm{IC}(c)} $$


The relation between the _FaITH_2010_ similarity and _Lin_1998_ similarity is:

$$ \mathrm{Sim}_\mathrm{FaITH} = \frac{\mathrm{Sim}_\mathrm{Lin}}{2 - \mathrm{Sim}_\mathrm{Lin}} $$


<pre class="r">
term_sim(dag, terms, method = "Sim_FaITH_2010")
</pre>

Paper link: https://doi.org/10.1007/978-3-642-17746-0_39.

## Sim_Relevance_2006

Note this is an annotation-based methods and the IC method is only restricted
to _IC_annotation_.

If thinking _Lin_1998_ is a measure of how close term $a$ and $b$ are to their
MICA term $c$, the relevance method corrects it by multiplying a factor which
considers the specificity of how $c$ brings the information. The factor is
calculated as $1-p(c)$ where $p(c)$ is the annotation-based probability $p(c)
= k/N$ where $k$ is the number of items annotated to $c$ and $N$ is the total
number of items annotated to the DAG. Then under the Relevance method, the
corrected IC of $c$ is:

$$ \mathrm{IC}_\mathrm{corrected}(c) = (1-p(c)) * \mathrm{IC}(c) $$

If using _Lin_1998_ as the similarity method, the corrected version Relevance
similarity is:

$$ 
\begin{align*}
\mathrm{Sim}(a, b) & = \frac{2*\mathrm{IC}_\mathrm{corrected}(c)}{\mathrm{IC}(a) + \mathrm{IC}(b)} \\
                   & = (1-p(c)) * \frac{2 * \mathrm{IC}(c)}{\mathrm{IC}(a) + \mathrm{IC}(b)} \\
                   & = (1-p(c)) * \mathrm{Sim}_\mathrm{Lin}(a, b)
\end{align*} $$


<pre class="r">
term_sim(dag, terms, method = "Sim_Relevance_2006")
</pre>

Paper link: https://doi.org/10.1186/1471-2105-7-302

## Sim_SimIC_2010

Note this is an annotation-based methods and the IC method is only restricted
to _IC_annotation_.

The _SimIC_2010_ method is an improved correction method of the Relevance
method because the latter works bad when $p(c)$ is very small. E.g., when
$1-p(c)$ is used as a correction factor, it cannot nicely distinguish $p(c) = 0.01$ and $p(c) = 0.001$ because for both $1 - p(c)$ are very close to 1.

The _SimIC_2010_ correction factor for MICA term $c$ is:


$$ \mathrm{IC}_\mathrm{corrected}(c) = \frac{\log(p(c))}{1 - \log(p(c))} * \mathrm{IC}(c) $$


Then the similarity (if using _Lin_1998_ as the original similarity method) is:

$$ \mathrm{Sim}(a, b) = \frac{\log(p(c))}{1 - \log(p(c))} * \mathrm{Sim}_\mathrm{Lin}(a, b) $$


<pre class="r">
term_sim(dag, terms, method = "Sim_SimIC_2010")
</pre>

Paper link: https://doi.org/10.48550/arXiv.1001.0958.


## Sim_XGraSM_2013

Note this is an annotation-based methods and the IC method is only restricted
to _IC_annotation_.

Being different from the _Relevance_ and _SimIC_2010_ methods that only use
the IC of the MICA term, the _XGraSM_2013_ method as well as the next two methods use
IC of a subset of common ancestor terms of $a$ and $b$, and it uses the mean
IC of them. The subset of common ancestor may have different names for
different methods.

_XGraSM_2013_ is the simplest one which uses informative common ancestors
(ICA) where IC of the common ancestor is positive.

$$ \mathrm{ICA}(a, b) = \{c \in \mathcal{A}_a^+ \cap \mathcal{A}_b^+: \mathrm{IC}(c) > 0\} $$

And mean IC among all ICA terms:

$$ \mathrm{IC}_\mathrm{mean} = \frac{1}{|\mathrm{ICA}(a, b)|} \sum_{\mathrm{t \in \mathrm{ICA}(a, b)}} \mathrm{IC}(t) $$

And applying _Lin_1998_ method, the semantic similarity is:

$$ \mathrm{Sim}(a, b) = 2 * \frac{\mathrm{IC}_\mathrm{mean}}{\mathrm{IC}(a) + \mathrm{IC}(b)} $$

<pre class="r">
term_sim(dag, terms, method = "Sim_XGraSM_2013")
</pre>

Paper link: https://doi.org/10.1186/1471-2105-14-284

## Sim_GraSM_2005

Note this is an annotation-based methods and the IC method is only restricted
to _IC_annotation_.

It applies a complicated way to selet a subset of common ancestors of terms $a$
and $b$. The main idea is that information transmission from ancestors to a
term may overlap and it aims to find paths from upstream that give exclusive
transmissions.

First, for a term $x$ define its disjunctive ancestors (DA) pairs $s$ and $t$
where there exists a path from $s$ to $x$ which does not pass $t$, and at the
same time there also exists a path from $t$ to $x$ which does not pass $s$. In
other words, the directed distance between $s$ and $t$ is zero. In this way,
the two $\mathrm{DA}(x)$ terms can transmit exclusive information to $x$.

Next, for every common ancestor $c$ of term $a$ and $b$, $c$ is only selected
if there is another common ancestor $c'$ which has smaller IC than $c$
($\mathrm{IC}(c) > \mathrm{IC}(c')$), an the pair $(c, c')$ is the DA of either
$a$ or $b$. The set of selected common ancestors is called the disjunctive
common ancestor (DCA).

And mean IC among all DCA terms:

$$ \mathrm{IC}_\mathrm{mean} = \frac{1}{|\mathrm{DCA}(a, b)|} \sum_{\mathrm{t \in \mathrm{DCA}(a, b)}} \mathrm{IC}(t) $$

And applying _Lin_1998_ method, the semantic similarity is:

$$ \mathrm{Sim}(a, b) = 2 * \frac{\mathrm{IC}_\mathrm{mean}}{\mathrm{IC}(a) + \mathrm{IC}(b)} $$

<pre class="r">
term_sim(dag, terms, method = "Sim_GraSM_2005")
</pre>

Paper link: https://doi.org/10.1145/1099554.1099658.

## Sim_EISI_2015

Note this is an annotation-based methods and the IC method is only restricted
to _IC_annotation_.

It also selects a specific subset of common ancestors of terms $a$ and $b$. It
only selects a common ancestor $c$ which can reach $a$ or $b$ via one of its
child terms that does not belong to the common ancestors (mutual exclusively
in $a$'s ancestors or in $b$'s ancestors). The set of the selected common
ancestors is called the exclusively inherited common ancestors (EICA).

$$ \mathrm{EICA}(a, b) = \{c \in \mathcal{A}_a \cap \mathcal{A}_b: \mathcal{C}_c \cap \left( (\mathcal{A}_a \cup \mathcal{A}_b) - (\mathcal{A}_a \cap \mathcal{A}_b) \neq \emptyset \right) \}  $$

And mean IC among all EICA terms:

$$ \mathrm{IC}_\mathrm{mean} = \frac{1}{|\mathrm{EICA}(a, b)|} \sum_{\mathrm{t \in \mathrm{EICA}(a, b)}} \mathrm{IC}(t) $$

And applying _Lin_1998_ method, the semantic similarit is:

$$ \mathrm{Sim}(a, b) = 2 * \frac{\mathrm{IC}_\mathrm{mean}}{\mathrm{IC}(a) + \mathrm{IC}(b)} $$

<pre class="r">
term_sim(dag, terms, method = "Sim_EISI_2015")
</pre>

Paper link: https://doi.org/10.1016/j.gene.2014.12.062

## Sim_AIC_2014

It uses the aggregate information content from ancestors. First define the
semantic weight denoted as $S_w$ of a term $t$ in the DAG:


$$ S_w(t) = \frac{1}{1 + \exp \left(-\frac{1}{\mathrm{IC}(t)} \right)} $$

Then the similarity is calculated as the fraction of aggegation from common
ancestors and the average aggregation from ancestors of $a$ and $b$ separately.

$$ \mathrm{Sim}(a, b) = \frac{2*\sum\limits_{t \in \mathcal{A}_a^+ \cap \mathcal{A}_b^+} S_w(t) }{ \sum\limits_{t \in \mathcal{A}_a^+} S_w(t) + \sum\limits_{t \in \mathcal{A}_b^+} S_w(t) }  $$

<pre class="r">
term_sim(dag, terms, method = "Sim_AIC_2014")
</pre>

Paper link: https://doi.org/10.1109/tcbb.2013.176.

## Sim_Zhang_2006

It uses the _IC_Zhang_2006_ IC method and uses _Lin_1998_ similarity method to calculate similarities:

$$ \mathrm{Sim}(a, b) = \frac{2*\mathrm{IC}_\mathrm{Zhang}(c)}{\mathrm{IC}_\mathrm{Zhang}(a) + \mathrm{IC}_\mathrm{Zhang}(b)} $$

<pre class="r">
term_sim(dag, terms, method = "Sim_Zhang_2006")
</pre>


## Sim_universal

It uses the _IC_universal_ IC method and uses the _Nunivers_ method to calculate similarities:

$$ \mathrm{Sim}(a, b) = \frac{2*\mathrm{IC}_\mathrm{Univers}(c)}{\max \{ \mathrm{IC}_\mathrm{Univers}(a), \mathrm{IC}_\mathrm{Univers}(b) \}} $$

<pre class="r">
term_sim(dag, terms, method = "Sim_universal")
</pre>


## Sim_Wang_2007

Similar as the _Sim_AIC_2014_ method, it is also aggregation from ancestors, but it uses the "S-value"
introduced in the _IC_Wang_2007_ sectionn in [4. Information content](v4_information_content.html).

$$ \mathrm{Sim}(a, b) = \frac{\sum\limits_{t \in \mathcal{A}_a^+ \cap \mathcal{A}_b^+} (S_a(t) + S_b(t)) }{ \sum\limits_{t \in \mathcal{A}_a^+} S_a(t) + \sum\limits_{t \in \mathcal{A}_b^+} S_b(t) }  $$


The contribution of different semantic relations can be set with the
`contribution_factor` parameter. The value should be a named numeric vector
where names should cover the relations defined in relations set in
`create_ontology_DAG()`. For example, if there are two relations "relation_a"
and "relation_b" set in the DAG, the value for contribution_factor can be set
as:

<pre class="r">
term_sim(dag, terms, method = "Sim_Wang_2007", 
    control = list(contribution_factor = c("relation_a" = 0.8, "relation_b" = 0.6)))
</pre>

By default 0.8 is set for "is_a" and 0.6 for "part_of".

If you are not sure what types of relations have been set, simply type the
`dag` object. The relation types will be printed there.

Paper link: https://doi.org/10.1093/bioinformatics/btm087.


# Edge-based methods

Methods introduced in this section replies on the distance between terms. Many
methods are defined originally based on the shortest distance between two
terms. This section extends them to also support their longest distance via
the LCA term.

## Sim_Rada_1989

It is based on the distance between term $a$ and $b$. It is defined as:

$$ \mathrm{Sim}(a, b) = \frac{1}{1 + D_\mathrm{sp}(a, b)} $$

which is based on the shortest distance between $a$ and $b$. Optionally,
the distance can also be the longest distance via the LCA term $c$.

$$ \mathrm{Sim}(a, b) = \frac{1}{1 + \mathrm{len}_c(a, b)} = \frac{1}{1 + \mathrm{len}(c, a) + \mathrm{len}(c, b)} $$

There is a parameter distance which takes value of `"longest_distances_via_LCA"`
(the default) or `"shortest_distances_via_NCA"`:

<pre class="r">
term_sim(dag, terms, method = "Sim_Rada_1989",
    control = list(distance = "shortest_distances_via_NCA"))
</pre>

Paper link: https://doi.org/10.1109/21.24528.


## Sim_Resnik_edge_2005

It is a normalized distance:

$$ \mathrm{Sim}(a, b) = 1 - \frac{D_\mathrm{sp}(a, b)}{2*\delta_\mathrm{max}} $$

where $2*\delta_\mathrm{max}$ can be thought as the possible maximal distance
between two terms in the DAG.

Similarly, the distance can also be the longest distance via LCA, then it is
consistent with the definition of $\delta_\mathrm{max}$ which are both based
on the longest distance.

$$ \mathrm{Sim}(a, b) = 1 - \frac{\mathrm{len}_c(a, b)}{2*\delta_\mathrm{max}} $$


There is a parameter distance which takes value of `"longest_distances_via_LCA"`
(the default) or `"shortest_distances_via_NCA"`:

<pre class="r">
term_sim(dag, terms, method = "Sim_Resnik_edge_2005",
    control = list(distance = "shortest_distances_via_NCA"))
</pre>

Paper link: https://doi.org/10.1145/1097047.1097051.


## Sim_Leocock_1998

It is similar as the _Sim_Resnik_edge_2005_ method, but it applies
log-transformation on the distance and the depth:

$$ \mathrm{Sim}(a, b) = 1 - \frac{\log(D_\mathrm{sp}(a, b))}{\log(2*\delta_\mathrm{max})} $$

where $2*\delta_\mathrm{max}$ can be thought as the possible maximal distance
between two terms in the DAG.

Similarly, the distance can also be the longest distance via LCA, then it is
consistent with the definition of $\delta_\mathrm{max}$ which are both based
on the longest distance.

$$ \mathrm{Sim}(a, b) = 1 - \frac{\log(\mathrm{len}_c(a, b))}{\log(2*\delta_\mathrm{max})} $$


There is a parameter distance which takes value of `"longest_distances_via_LCA"`
(the default) or `"shortest_distances_via_NCA"`:

<pre class="r">
term_sim(dag, terms, method = "Sim_Leocock_1998",
    control = list(distance = "shortest_distances_via_NCA"))
</pre>

Paper link: https://ieeexplore.ieee.org/document/6287675.


## Sim_WP_1994

It is based on the depth of the LCA term $c$ and the longest distance between
term $a$ and $b$ via $c$:

$$
\begin{align*}
 \mathrm{Sim}(a, b) & = \frac{2*\delta(c)}{\mathrm{len}(c, a) + \mathrm{len}(c, b) + 2*\delta(c)} \\
   & =  \frac{2*\delta(c)}{\mathrm{len}_c(a, b) + 2*\delta(c)}
\end{align*}
$$

And it can also be written in the *Lin_1998* form:

$$
\begin{align*}
 \mathrm{Sim}(a, b) & = \frac{2*\delta(c)}{\delta(c) + \mathrm{len}(c, a) + \delta(c) + \mathrm{len}(c, b)} \\
   & =  \frac{2*\delta(c)}{\delta_c(a) + \delta_c(b)}
\end{align*}
$$

where in the denominator are the depths of $a$ and $b$ via $c$.

<pre class="r">
term_sim(dag, terms, method = "Sim_WP_1994")
</pre>

Paper link: https://doi.org/10.3115/981732.981751.

## Sim_Slimani_2006

It is a correction of the *Sim_WP_1994* method. The correction factor for term $a$ and $b$ regarding to their LCA term $c$ is:

$$ \mathrm{Sim}(a, b) = \mathrm{CF}(a, b) * \mathrm{Sim}_\mathrm{WP}(a, b) $$

The correction factor $\mathrm{CF}(a, b)$ is based whether $a$ and $b$ are in
ancestor/offspring relationship or not.

$$ \mathrm{CF}(a, b) = \left\{ \begin{array}{ll}
\min\{ \delta(a), \delta(b)\} - \delta(c) = \min\{\mathrm{len}(c, a), \mathrm{len}(c, b)\} & \textit{a} \text{ and } \textit{b} \text{ are not ancestor-offspring} \\
\frac{1}{1 + |\delta(a) - \delta(b)|} = \frac{1}{1 + \mathrm{len}(a,b)} & \textit{a} \text{ and } \textit{b} \text{ are ancestor-offspring}
\end{array} \right. 
$$

<pre class="r">
term_sim(dag, terms, method = "Sim_Slimani_2006")
</pre>

Paper link: https://zenodo.org/record/1075130.

## Sim_Shenoy_2012

It is also a correction of the *Sim_WP_1994* method. The correction factor for term $a$ and $b$ is:

$$  \mathrm{CF}(a, b) = \left\{ \begin{array}{ll}
1 & \textit{a} \text{ and } \textit{b} \text{ are not ancestor-offspring} \\
\exp(-\frac{D_\mathrm{sp}(a, b)}{\delta_\mathrm{max}})) & \textit{a} \text{ and } \textit{b} \text{ are ancestor-offspring}
\end{array} \right. $$

$D_\mathrm{sp}$ can be replaced with $\mathrm{len}(a, b)$ if the longest distance is used.


There is a parameter distance which takes value of
`"longest_distances_via_LCA"` (the default) or `"shortest_distances_via_NCA"`:

<pre class="r">
term_sim(dag, terms, method = "Sim_Shenoy_2012",
    control = list(distance = "shortest_distances_via_NCA"))
</pre>

Paper link: https://doi.org/10.48550/arXiv.1211.4709.


## Sim_Pekar_2002

It is very similar to the _Sim_WP_1994_ method:

$$
\begin{align*}
\mathrm{Sim}(a, b) &= \frac{\delta(c)}{\mathrm{len}(c, a) + \mathrm{len}(c, b) + \delta(c)} \\
                   &= \frac{\delta(c)}{\delta(c) + \mathrm{len}(c, a) + \delta(c) + \mathrm{len}(c, b) - \delta(c)} \\
                   &= \frac{\delta(c)}{\delta_c(a) + \delta_c(b) - \delta(c)}
\end{align*}
$$


And the relationship to $\mathrm{Sim}_\mathrm{WP}$ is:

$$ \mathrm{Sim}_\mathrm{Pekar}(a, b) = \frac{\mathrm{Sim}_\mathrm{WP}(a, b)}{2 - \mathrm{Sim}_\mathrm{WP}(a, b)} $$


<pre class="r">
term_sim(dag, terms, method = "Sim_Pekar_2002")
</pre>

Paper link: https://aclanthology.org/C02-1090/.

## Sim_Stojanovic_2001

It is purely based on the depth of term $a$, $b$ and their LCA term $c$.


$$ \mathrm{Sim}(a, b) = \frac{\delta(c)}{\delta(a) + \delta(b) - \delta(c)} $$

The similarity value might be negative because there is no restrction that the
path from root to $a$ or $b$ must pass $c$.


<pre class="r">
term_sim(dag, terms, method = "Sim_Stojanovic_2001")
</pre>

Paper link: https://doi.org/10.1145/500737.500762.


## Sim_Wang_edge_2012

It is calculated as:

$$ 
\begin{align*}
\mathrm{Sim}(a, b) & = \frac{\mathrm{len}(r, c)^2}{\mathrm{len}_c(r, a)*\mathrm{len}_c(r, b)} \\ 
   & = \frac{\delta(c)^2}{\delta_c(a)*\delta_c(b)}
\end{align*}
$$

where $r$ is the root term.


<pre class="r">
term_sim(dag, terms, method = "Sim_Wang_edge_2012")
</pre>

Paper link: https://doi.org/10.1186/1477-5956-10-s1-s18.


## Sim_Zhong_2002

For a term $x$, it first calculates a "mile-stone" value based on the depth as

$$ m(x) = 2^{-\delta(x) - 1} $$

The the distance bewteen term $a$ and $b$ via LCA term $c$ is:

$$ 
\begin{align*}
D(a, b) & = D(c, a) + D(c, b) \\
   & = m(c) - m(a) + m(c) + m(b) \\
   & = 2^{-\delta(c)} - 2^{-\delta(a) - 1} - 2^{-\delta(b) - 1}
\end{align*}
$$

We can change original $\delta(a)$ and $\delta(b)$ to $\delta_c(a)$
and $\delta_c(b)$ to require that the depth to reach $a$ and $b$ should go through $c$. Then above
equation becomes


$$ 
\begin{align*}
D(a, b) & = 2^{-\delta(c)} - 2^{-\delta_c(a) - 1} - 2^{-\delta_c(b) - 1} \\
    & = 2^{-\delta(c)} - 2^{-\delta(c)-\mathrm{len}(c,a)-1} - 2^{-\delta(c)-\mathrm{len}(c,b)-1} \\
    & = 2^{-\delta(c)} \left( 1 - 2^{-\mathrm{len}(c,a)-1} - 2^{-\mathrm{len}(c,b)-1} \right)
\end{align*}
$$

Then when $a = b$ (the two terms are identical), $D(a, b) = 0$ and when $c = r$ (common ancestor only includes root) and $\mathrm{len}(r, a) \to \infty$, $\mathrm{len}(r, b) \to \infty$ (root has infinite distance to the terms), $D(a, b)$ reaches maximal of 1. So the similarity

$$ \mathrm{Sim}(a, b) = 1 - D(a, b)$$

ranges between 0 and 1.


<pre class="r">
term_sim(dag, terms, method = "Sim_Zhong_2002")
</pre>

Paper link: https://doi.org/10.1007/3-540-45483-7_8.


## Sim_AlMubaid_2006

It also takes accout of the distance between term $a$ and $b$, as well as the
depth of the LCA term $c$ in the DAG. The distance is calculated as:

$$ D(a, b) = \log(1 + D_\mathrm{sp}(a, b)*(\sigma_\mathrm{max} - \sigma(c))) $$

To scale $D(a, b)$ into the range of `[0, 1]`, we can calculate the smallest value as zero when $a = b$.
$D(a, b)$ reaches maximal when $D_\mathrm{sp}(a, b)$ reach possible maximal which is $2*\delta_\mathrm{max}$. Then we can define the maximal of $D(a, b)$ as

$$ D_\mathrm{max} = \log(1 + 2*\delta_\mathrm{max} * \delta_\mathrm{max}) $$

And the similarity is:

$$ \mathrm{Sim}(a, b) = 1 - D(a, b)/D_\mathrm{max} $$


There is a parameter distance which takes value of
`"longest_distances_via_LCA"` (the default) or `"shortest_distances_via_CA"`:

<pre class="r">
term_sim(dag, terms, method = "Sim_AlMubaid_2006",
    control = list(distance = "shortest_distances_via_CA"))
</pre>

Paper link: https://doi.org/10.1109/IEMBS.2006.259235.


## Sim_Li_2003

It is similar to the *Sim_AlMubaid_2006* method, but uses a non-linear form:

$$ \mathrm{Sim}(a, b) = \exp(-0.2*D_\mathrm{sp}(a, b)) * \tanh(0.6*\delta(c)) $$


There is a parameter distance which takes value of `"longest_distances_via_LCA"`
(the default) or `"shortest_distances_via_CA"`:

<pre class="r">
term_sim(dag, terms, method = "Sim_Li_2003",
    control = list(distance = "shortest_distances_via_CA"))
</pre>

Paper link: https://doi.org/10.1109/TKDE.2003.1209005.


# Hybrid methods

Hybrid methods use both DAG structure information and IC.

## Sim_RSS_2013

The similarity is adjusted by the positions of term $a$, $b$ and the LCA term $c$ in the DAG. The similarity is defined as:

$$ \mathrm{Sim}(a, b) = \frac{\delta_\mathrm{max}}{\delta_\mathrm{max} + D_\mathrm{sp}(a, b)} * \frac{\alpha}{\alpha + \beta} $$

where $D_\mathrm{sp}(a, b)$ can also be the longest distance via LCA. $\alpha$ and $\beta$ in the second term are defined as:

$$
\begin{align*}
\alpha & = \delta(c) \\
\beta & = \min\{ \eta(a), \eta(b) \}
\end{align*}
$$

where $\alpha$ is the depth of LCA, $\beta$ corresponds to the distance to
leaves, which is the smaller height of $a$ and $b$ in the DAG.


There is a parameter distance which takes value of
`"longest_distances_via_LCA"` (the default) or `"shortest_distances_via_CA"`:

<pre class="r">
term_sim(dag, terms, method = "Sim_RSS_2013",
    control = list(distance = "shortest_distances_via_CA"))
</pre>

Paper link: https://doi.org/10.1371/journal.pone.0066745.


## Sim_HRSS_2013

It is similar to the *Sim_RSS_2013* method, but it uses information content
instead of the distance to adjust the similarity.

It first defines the semantic distance between term $a$ and $b$ as the sum of
the distance to their MICA term $c$:

$$ D(a, b) = D(c, a) + D(c, b) $$

And the distance between an ancestor to a term is:

$$ 
\begin{align*}
D(c, a) & = \mathrm{IC}(a) - \mathrm{IC}(c) \\
D(a, b) & = D(c, a) + D(c, b) = \mathrm{IC}(a) + \mathrm{IC}(b) - 2*\mathrm{IC}(c)
\end{align*}
$$

Similarly, the similarity is also corrected by the position of MICA term and $a$, $b$ in the DAG:

$$ \mathrm{Sim}(a, b) = \frac{1}{1 + D(a, b)} * \frac{\alpha}{\alpha + \beta} $$

where

$$ \alpha = \mathrm{IC}(c) $$

And beta is the average of the maximal semantic distance of $a$ and $b$ to leaves.

$$ \beta = \frac{D(a, l_a) + D(b, l_b)}{2} = \frac{\mathrm{IC}(l_a) - \mathrm{IC}(a) + \mathrm{IC}(l_b) - \mathrm{IC}(b)}{2} $$

where $l_a$ or $l_b$ is the leaf with the highest IC that $a$ or $b$ can reach (i.e. the most
informative leaf)

$$ \mathrm{IC}(l_a) = \max_{z \in \mathcal{L}(a)} \mathrm{IC}(z) $$


<pre class="r">
term_sim(dag, terms, method = "Sim_HRSS_2013")
</pre>

Paper link: https://doi.org/10.1371/journal.pone.0066745.


## Sim_Shen_2010

It is based on the information content of terms on the path connecting term $a$
and $b$ via their MICA term $c$.

Denote a list of terms `a, ..., c, ..., b` which are composed by the shortest
path from $c$ to $a$ and from $c$ to $b$, the distance between $a$ and $b$
is the sum of $1/\mathrm{IC}$ of the terms on the path. Denote $L_c(a, b)$ as the set of terms on the shortest path connecting $a$ and
$b$ via the MICA term $c$, the similarity is:

$$ \mathrm{Sim}(a, b) = 1 - \frac{\arctan \left( \sum\limits_{x \in L_c(a, b)} \frac{1}{\mathrm{IC}(x)} \right)}{\pi/2} $$


<pre class="r">
term_sim(dag, terms, method = "Sim_Shen_2010")
</pre>

Paper link: https://doi.org/10.1109/BIBM.2010.5706623.


## Sim_SSDD_2013

It is similar to the _Sim_Shen_2010_ method which also sums information along the
path passing through the LCA term. Instead of summing the information
contents, the _Sim_SSDD_2013_ method sums up a so-called "T-value" which relies on the
DAG structure.

Denote $L_c(a, b)$ as the set of terms on the shortest path connecting $a$ and $b$ via the LCA term $c$,
the similarity is calculated as:

$$ \mathrm{Sim}(a, b) = 1 - \frac{\arctan \left( \sum\limits_{x \in L_c(a, b)} T(x) \right) }{\pi/2} $$

The T-value $T(x)$ depends on the DAG structure which considers both parents and children of $x$. The
definition of $T(x)$ is:

$$
T(x) = \left\{ \begin{array}{ll}
1 & \text{if }\textit{x}\text{ is a root} \\
\frac{1}{|\mathcal{P}_x|} \sum\limits_{t \in \mathcal{P}_x}(w * T(t)) & \text{otherwise}
\end{array} \right.
$$

which means T-value of a term is an average of the weighted T-values of its parents. The
weight $w$ measures the fraction of information a parent $t$ transmitting to downstream of the DAG via $x$, defined as:

$$ w = \frac{|D_x^+|}{|D_t^+|} $$

$w \leq 1$ as all offsprings of $x$ are also offspring of its parent $t$.


<pre class="r">
term_sim(dag, terms, method = "Sim_SSDD_2013")
</pre>

Paper link: https://doi.org/10.1016/j.ygeno.2013.04.010.


## Sim_Jiang_1997

First semantic distance between term $a$ and $b$ via MICA term $c$ is defined as:

$$ D(a, b) = \mathrm{IC}(a) + \mathrm{IC}(b) - 2*\mathrm{IC}(c) $$ 

Then there are several normalization methods to change the distance to similarity and to scale it into the range of `[0, 1]`.

- `"max"`: $1 - \frac{D(a, b)}{2*\mathrm{IC}_\mathrm{max}}$
- `"Couto"`: $\min\{ 1, \frac{D(a, b)}{\mathrm{IC}_\mathrm{max}} \}$
- `"Lin"`: $1 - \frac{D(a, b)}{\mathrm{IC}(a) + \mathrm{IC}(b)}$ which is the same as the *Sim_Lin_1998* method
- `"Garla"`: $1 - \frac{\log(D(a, b) + 1)}{\log(2*\mathrm{IC}_\mathrm{max} + 1)}$
- `"log-Lin"`: $1 - \frac{\log(D(a, b) + 1)}{\log(\mathrm{IC}(a) + \mathrm{IC}(b) + 1)}$
- `"Rada"`: $\frac{1}{1 + D(a, b)}$

The normalization methods can be set via the parameter `norm_method`:

<pre class="r">
term_sim(dag, terms, method = "Sim_Jiang_1997", control = list(norm_method = "max"))
term_sim(dag, terms, method = "Sim_Jiang_1997", control = list(norm_method = "Couto"))
term_sim(dag, terms, method = "Sim_Jiang_1997", control = list(norm_method = "Lin"))
term_sim(dag, terms, method = "Sim_Jiang_1997", control = list(norm_method = "Garla"))
term_sim(dag, terms, method = "Sim_Jiang_1997", control = list(norm_method = "log-Lin"))
term_sim(dag, terms, method = "Sim_Jiang_1997", control = list(norm_method = "Rada"))
</pre>

Paper link: https://aclanthology.org/O97-1002/.


# Annotation-count based methods

Denote $A$ and $B$ as the sets of items annotated to term $a$ and $b$, and $U$ as the universe set
of all items annotated to the DAG.

## Sim_Kappa

The definition of kappa coeffient is a little bit complex. First let's format the two sets
into a contigency table:

<style>
table.contigency-table {
    border-collapse: collapse;
}
table.contigency-table td {
    border: 1px solid black;
    padding: 2px 5px;
}
.symbol {
    font-family: Times;
    font-style: italic;
}
</style>
<center>
<table class="contigency-table">
    <tr><td rowspan='2' colspan='2'></td><td colspan='2'>In set <b><i>B</i></b></td></tr>
    <tr><td>Yes</td><td>No</td></tr>
    <tr><td rowspan='2'>In set <b><i>A</i></b></td><td>Yes</td><td class="symbol">a</td><td class="symbol">b</td></tr>
    <tr><td>No</td><td class="symbol">c</td><td class="symbol">d</td></tr>
</table>
</center>

where $a$, $b$, $c$, $d$ are the numbers of items that fall in each category.

Let's calculate $p_\mathrm{obs}$ (probability of observed agreement, both yes or both no) and $p_\mathrm{exp}$ (probability of expected agreement) as: 

$$
\begin{align*}
p_\mathrm{obs} & = \frac{a+d}{a+b+c+d} \\
p_\mathrm{Yes} & = \frac{a+b}{a+b+c+d} * \frac{a+c}{a+b+c+d} \\
p_\mathrm{No} & = \frac{c+d}{a+b+c+d} * \frac{b+d}{a+b+c+d} \\
p_\mathrm{exp} & = p_\mathrm{Yes} + p_\mathrm{No}
\end{align*}
$$

where $p_\mathrm{obs}$ is the probability of an item in both sets or neither in both sets,
$p_\mathrm{Yes}$ is the probability of an item in both sets by random (by assuming the events of an item in set $A$ and set $B$ are independent), $p_\mathrm{No}$ is the probability of an item
not in the two sets by random, and $p_\mathrm{exp}$ is the probability of an item either both in 
the two sets or not in the two sets by random.

The kappa coeffcient is calculated as:

$$ \mathrm{Sim}(a, b) = \mathrm{Kappa}(a, b) = \frac{p_\mathrm{obs} - p_\mathrm{exp}}{1 - p_\mathrm{exp}}$$

Note the Kappa coeffcient is possible to be negative.

The universe set can be set via the parameter `anno_universe`. By default it
is the total items annotated to the whole DAG.

<pre class="r">
term_sim(dag, terms, method = "Sim_kappa",
    control = list(anno_universe = ...))
</pre>

## Sim_Jaccard, Sim_Dice and Sim_Overlap

Definitions of the Jaccard, Dice and overlap coeffcients are similar. 
The Jaccard coeffcient is:

$$ \mathrm{Jaccard}(a, b) = \frac{|A \cap B|}{|A \cup B|} $$

The Dice coeffcient is:

$$ \mathrm{Dice}(a, b) = \frac{2*|A \cap B|}{|A| + |B|} $$

The overlap coeffcient is:

$$ \mathrm{Overlap}(a, b) = \frac{|A \cap B|}{\min\{|A|, |B|\}} $$


Dice and Jaccard coeffcients have a relation of:

$$ \mathrm{Jaccard} = \frac{\mathrm{Dice}}{2 - \mathrm{Dice}} $$

The universe set can be set via the parameter `anno_universe`. By default it
is the total items annotated to the whole DAG.

<pre class="r">
term_sim(dag, terms, method = "Sim_Jaccard",
    control = list(anno_universe = ...))
term_sim(dag, terms, method = "Sim_Dice",
    control = list(anno_universe = ...))
term_sim(dag, terms, method = "Sim_Overlap",
    control = list(anno_universe = ...))
</pre>

# Session Info

```{r}
sessionInfo()
```

