---
title: "Similarity between two groups of terms"
author: "Zuguang Gu ( z.gu@dkfz.de )"
date: '`r Sys.Date()`'
output: 
  html_vignette:
    css: main.css
    toc: true
vignette: >
  %\VignetteIndexEntry{6. Similarity between two groups of terms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE, message = FALSE}
library(knitr)
knitr::opts_chunk$set(
    error = FALSE,
    tidy  = FALSE,
    message = FALSE,
    warning = FALSE,
    fig.align = "center")
```



The methods of group similarity implemented in **simona** are mainly from
the [supplementary file](https://academic.oup.com/bib/article/18/5/886/2562801#supplementary-data)
of the paper ["Mazandu et al., Gene Ontology semantic similarity tools: survey
on features and challenges for biological knowledge discovery. Briefings in
Bioinformatics 2017"](https://doi.org/10.1093/bib/bbw067). Original
denotations have been slightly modified to make them more consistent. Also
more explanations have been added in this vignette. 


There are two groups of terms denoted as $T_p$ and $T_q$ represented as two sets:

$$ T_p = \{ a_1, a_2, ...\} \\
   T_q = \{ b_1, b_2, ... \} $$

where $a_i$ is a term in set $T_p$ and $b_j$ is a term in set $T_q$.


The wrapper function `group_sim()` calculates semantic similarities between two groups
of terms with a specific method. Note the method name can be partially matched.


<pre class="r">
group_sim(dag, group1, group2, method = ..., control = list(...))
</pre>

Some of the group similarity methods have no assumption of which similarity
measure between single terms to use. If there are annotation already provided
in the DAG object, by default *Sim_Lin_1998* is used, or else *Sim_WP_1994* is
used. The term similarity method can be set via the `term_sim_method`
parameter in `control`. Additionally parameters for a specific `term_sim_method`
can also be set in `control`.

<pre class="r">
group_sim(dag, group1, group2, method = ..., 
    control = list(term_sim_method = ...))
</pre>

# Pairwise term similarity-based methods

## GroupSim_pairwise_avg

Denote $S(a, b)$ as the semantic similarity between term $a$ and $b$ where $a$
is from group $p$ and $b$ is from group $q$, The similarity between group $p$
and group $q$ is the average similarity of every pair of individual terms in
the two groups:

$$ \mathrm{GroupSim}(p, q) = \frac{1}{|T_p|*|T_q|} \sum_{a \in T_p, b \in T_q}S(a, b) $$

<pre class="r">
group_sim(dag, group1, group2, method = "GroupSim_pairwise_avg",
    control = list(term_sim_method = ...))
</pre>

## GroupSim_pairwise_max

The similarity is defined as the maximal $S(a, b)$ among all pairs of terms in group $p$ and $q$:

$$ \mathrm{GroupSim}(p, q) = \max_{a \in T_p, b \in T_q}S(a, b) $$


<pre class="r">
group_sim(dag, group1, group2, method = "GroupSim_pairwise_max",
    control = list(term_sim_method = ...))
</pre>

## GroupSim_pairwise_BMA

BMA stands for "best-match average". First define similarity of a term $x$ to a group of terms $T$ as

$$ S(x, T) = \max_{y \in T} S(x, y) $$

which corresponds to the most similar term in $T$ to $x$. Then the BMA similarity is calculated as:

$$ \mathrm{GroupSim}(p, q) = \frac{1}{2}\left( \frac{1}{|T_p|}\sum_{a \in T_p} S(a, T_q) + \frac{1}{|T_q|}\sum_{b \in T_q} S(b, T_p) \right) $$

<pre class="r">
group_sim(dag, group1, group2, method = "GroupSim_pairwise_BMA",
    control = list(term_sim_method = ...))
</pre>

## GroupSim_pairwise_BMM

BMM stands for "best-match max". It is defined as:

$$ \mathrm{GroupSim}(p, q) = \max \left \{ \frac{1}{|T_p|}\sum_{a \in T_p} S(a, T_q), \frac{1}{|T_q|}\sum_{b \in T_q} S(b, T_p) \right \} $$


<pre class="r">
group_sim(dag, group1, group2, method = "GroupSim_pairwise_BMM",
    control = list(term_sim_method = ...))
</pre>

Paper link: https://doi.org/10.1093/bioinformatics/btq064.

## GroupSim_pairwise_ABM

ABM stands for "average best-match". It is defined as:

$$ \mathrm{GroupSim}(p, q) = \frac{1}{|T_q| + |T_q|} \left( \sum_{a \in T_p} S(a, T_q) + \sum_{b \in T_q} S(b, T_p) \right)  $$

<pre class="r">
group_sim(dag, group1, group2, method = "GroupSim_pairwise_ABM",
    control = list(term_sim_method = ...))
</pre>

Paper link: https://doi.org/10.1186/1471-2105-14-284.

## GroupSim_pairwise_HDF

First define the distance of a term $x$ to a group of terms $T$:

$$D(x, T) = 1 - S(x, T)$$

Then the Hausdorff distance between two groups are:

$$ \mathrm{HDF}(p, q) = \max \left\{ \max_{a \in T_p} D(a, T_q), \max_{b \in T_q} D(b, T_q) \right\} $$


This final similarity is:

$$ \mathrm{GroupSim}(p, q) = 1 - \mathrm{HDF}(p, q) $$


<pre class="r">
group_sim(dag, group1, group2, method = "GroupSim_pairwise_HDF",
    control = list(term_sim_method = ...))
</pre>


## GroupSim_pairwise_MHDF

Instead of using the maximal distance from a group to the other group, MHDF uses mean distance:


$$ \mathrm{MHDF}(p, q) = \max \left\{ \frac{1}{|T_p|} \sum_{a \in T_p} D(a, T_q), \frac{1}{|T_q|} \sum_{b \in T_q} D(b, T_q) \right\} $$

This final similarity is:

$$ \mathrm{GroupSim}(p, q) = 1 - \mathrm{MHDF}(p, q) $$


<pre class="r">
group_sim(dag, group1, group2, method = "GroupSim_pairwise_MHDF",
    control = list(term_sim_method = ...))
</pre>

Paper link: https://doi.org/10.1109/ICPR.1994.576361.

## GroupSim_pairwise_VHDF

It is defined as:


$$ \mathrm{VHDF}(p, q) = \frac{1}{2} \left( \sqrt{\frac{1}{|T_p|} \sum_{a \in T_p} D^2(a, T_q)} + \sqrt{\frac{1}{|T_q|} \sum_{b \in T_q} D^2(b, T_q)} \right) $$


This final similarity is:

$$ \mathrm{GroupSim}(p, q) = 1 - \mathrm{VHDF}(p, q) $$

<pre class="r">
group_sim(dag, group1, group2, method = "GroupSim_pairwise_VHDF",
    control = list(term_sim_method = ...))
</pre>

Paper link: https://doi.org/10.1073/pnas.0702965104.

## GroupSim_pairwise_Froehlich_2007

The similarity is:


$$ \mathrm{GroupSim}(p, q) = \exp(-\mathrm{HDF}(p, q)) $$


<pre class="r">
group_sim(dag, group1, group2, method = "GroupSim_pairwise_Froehlich_2007",
    control = list(term_sim_method = ...))
</pre>

Paper link: https://doi.org/10.1186/1471-2105-8-166.


## GroupSim_pairwise_Joeng_2014

Similar to VHDF, but it directly uses the similarity:

$$ \mathrm{GroupSim}(p, q) = \frac{1}{2} \left( \sqrt{\frac{1}{|T_p|} \sum_{a \in T_p} S^2(a, T_q)} + \sqrt{\frac{1}{|T_q|} \sum_{b \in T_q} S^2(b, T_q)} \right) $$


<pre class="r">
group_sim(dag, group1, group2, method = "GroupSim_pairwise_Joeng_2014",
    control = list(term_sim_method = ...))
</pre>

Paper link: https://doi.org/10.1109/tcbb.2014.2343963.

# Pairwise edge-based methods

## GroupSim_SimALN

It is based on the average distance between every pair of terms in the two groups:

$$ \mathrm{GroupSim}(p, q) = \exp\left(-\frac{1}{|T_p|*|T_q|} \sum_{a \in T_p, b \in T_q} D_\mathrm{sp}(a, b)\right) $$

Or use the longest distance between two terms:

$$ \mathrm{GroupSim}(p, q) = \exp\left(-\frac{1}{|T_p|*|T_q|} \sum_{a \in T_p, b \in T_q} \mathrm{len}(a, b)\right) $$


There is a parameter distance which takes value of `"longest_distances_via_LCA"`
(the default) or `"shortest_distances_via_NCA"`:


<pre class="r">
group_sim(dag, group1, group2, method = "GroupSim_SimALN",
    control = list(distance = "shortest_distances_via_NCA"))
</pre>

Paper link: https://doi.org/10.1109/CBMS.2008.27.


# Groupwise IC-based methods

This category of methods depend on the IC of terms in the two groups as well as their ancestor terms.

## GroupSim_SimGIC, GroupSim_SimDIC, GroupSim_SimUIC, GroupSim_SimUI, GroupSim_SimDB, GroupSim_SimUB and GroupSim_SimNTO

Denote $A$ and $B$ as the two sets of ancestors of terms in group $p$ and $q$ respectively:


$$
\begin{align*}
 \mathcal{A}_p &= \bigcup_{a \in T_p} \mathcal{A}_a \\
 \mathcal{A}_q &= \bigcup_{b \in T_q} \mathcal{A}_b \\
\end{align*}
$$

The _GroupSim_SimGIC_, _GroupSim_SimDIC_, _GroupSim_SimUIC_, _GroupSim_SimUI_,
_GroupSim_SimDB_,  _GroupSim_SimUB_ and _GroupSim_SimNTO_ are very similar.
They are defined as:


$$
\begin{align*}
 \mathrm{GroupSim}_\mathrm{SimGIC}(p, q) &= \frac{\sum\limits_{x \in \mathcal{A}_p \cap \mathcal{A}_q} \mathrm{IC}(x)}{\sum\limits_{x \in \mathcal{A}_p \cup \mathcal{A}_q} \mathrm{IC}(x)} \\
 \mathrm{GroupSim}_\mathrm{SimDIC}(p, q) &= \frac{2 * \sum\limits_{x \in \mathcal{A}_p \cap \mathcal{A}_q} \mathrm{IC}(x)}{\sum\limits_{x \in \mathcal{A}_p} \mathrm{IC}(x) + \sum\limits_{x \in \mathcal{A}_q} \mathrm{IC}(x)} \\
 \mathrm{GroupSim}_\mathrm{SimUIC}(p, q) &= \frac{2 * \sum\limits_{x \in \mathcal{A}_p \cap \mathcal{A}_q} \mathrm{IC}(x)}{\max\left\{\sum\limits_{x \in \mathcal{A}_p} \mathrm{IC}(x), \sum\limits_{x \in \mathcal{A}_q} \mathrm{IC}(x) \right\}}  \\
 \mathrm{GroupSim}_\mathrm{SimUI}(p, q) &= \frac{|\mathcal{A}_p \cap \mathcal{A}_q|}{|\mathcal{A}_p \cup \mathcal{A}_q|} \\
 \mathrm{GroupSim}_\mathrm{SimDB}(p, q) &= \frac{2*|\mathcal{A}_p \cap \mathcal{A}_q|}{|\mathcal{A}_p| +  |\mathcal{A}_q|} \\
 \mathrm{GroupSim}_\mathrm{SimUB}(p, q) &= \frac{|\mathcal{A}_p \cap \mathcal{A}_q|}{\max\{|\mathcal{A}_p|,  |\mathcal{A}_q|\}} \\
 \mathrm{GroupSim}_\mathrm{SimNTO}(p, q) &= \frac{|\mathcal{A}_p \cap \mathcal{A}_q|}{\min\{|\mathcal{A}_p|,  |\mathcal{A}_q|\}}
\end{align*}
$$

IC method can be set via the `control` argument. By default if there is annotation associated, _IC_annotation_ is used, or else _IC_offspring_ is used.

<pre class="r">
group_sim(dag, group1, group2, method = "GroupSim_SimGIC",
    control = list(IC_method = ...))
</pre>

## GroupSim_SimCOU

Let's write $\mathcal{A}_p$ and $\mathcal{A}_q$ as two vectors $\mathbf{v_p}$
and $\mathbf{v_q}$. Taking $\mathr{v_p}$ as an example, it is $\mathbf{v_p} = (w_1, ..., w_n)$ where $n$
is the number of total terms in the DAG. The value $w_i$ is assigned to the corresponding term $t_i$
and is defined as:

$$
\mathcal{w}_{i} = \left\{ \begin{array}{ll}
\mathrm{IC}(t_i) & \textrm{if} t_i \in \mathcal{A}_p \\
0 & \textrm{otherwise}
\end{array} \right.
$$

The semantic similarity is defined as the cosine similarity between the two vectors:

$$ \mathrm{GroupSim}(a, b) = \frac{ \mathbf{v_p} \cdot \mathbf{v_q} }{\left \| \mathbf{v_p} \right \| \cdot  \left \| \mathbf{v_q} \right \|} $$

It can also be written as:

$$ \mathrm{GroupSim}(a, b) = \frac{\sum\limits_{x \in \mathcal{A}_p \cap \mathcal{A}_q}\mathrm{IC}(x)^2}{\sqrt{\sum\limits_{x \in \mathcal{A}_p}\mathrm{IC}(x)^2} \cdot \sqrt{\sum\limits_{x \in \mathcal{A}_q}\mathrm{IC}(x)^2}} $$

## GroupSim_SimCOT

The semantic similarity is defined as:

$$
\begin{align*}
 \mathrm{GroupSim}(a, b) &= \frac{ \mathbf{v_p} \cdot \mathbf{v_q} }{\left \| \mathbf{v_p} \right \|^2 +  \left \| \mathbf{v_q} \right \|^2 - \mathbf{v_p} \cdot \mathbf{v_q}} \\
     &=  \frac{\sum\limits_{x \in \mathcal{A}_p \cap \mathcal{A}_q}\mathrm{IC}(x)^2}{\sum\limits_{x \in \mathcal{A}_p \cup \mathcal{A}_q}\mathrm{IC}(x)^2}
\end{align*}
$$



# Groupwise edge-based methods

## GroupSim_SimLP

It is the largest depth of terms in $\mathcal{A}_p \cap \mathcal{A}_q$.

$$ \mathrm{GroupSim}(p, q) = \max\{\delta(t): t \in \mathcal{A}_p \cap \mathcal{A}_q\} $$


<pre class="r">
group_sim(dag, group1, group2, method = "GroupSim_SimLP")
</pre>


Link: https://bioconductor.org/packages/release/bioc/vignettes/GOstats/inst/doc/GOvis.html#go-induced-distances.


## GroupSim_Ye_2005

It is a normalized version of *GroupSim_SimLP*:

$$
\begin{align*}
\mathrm{GroupSim}(p, q) &= \max\left\{\frac{\delta(t) - \delta_\mathrm{min}}{\delta_\mathrm{max} - \delta_\mathrm{min}}: t \in \mathcal{A}_p \cap \mathcal{A}_q\right\} \\
    &= \max\left\{\frac{\delta(t) }{\delta_\mathrm{max}}: t \in \mathcal{A}_p \cap \mathcal{A}_q\right\}
\end{align*}
$$

Since the minimal depth is zero for root. 


<pre class="r">
group_sim(dag, group1, group2, method = "GroupSim_Ye_2005")
</pre>

Paper link: https://doi.org/10.1038/msb4100034.

# Annotated items-based methods

This category of methods consider the items annotated to the two groups of terms. 


## GroupSim_SimCHO

It is based on the annotated items. Denote $\sigma(t)$ as the total number of
annotated items of $t$ (after merging all its offspring terms). The similarity
is calculated as:

$$
\mathrm{GroupSim}(p, q) = \frac{\log(C_{pq})}{\log(C_\mathrm{min}/C_\mathrm{max})}
$$

where $C_{pq} = \min\{\sigma(t): t \in T_p \cap T_q \}$, $C_\mathrm{min}$ is the minimal number of annotated items in the DAG which in most cases is 1, $C_\mathrm{max}$ is the maximal number of annotated items, which is the total number of items annotated to the complete DAG.

The similarity can also be written in form of $\mathrm{IC}_\mathrm{anno}$:

$$
\mathrm{GroupSim}(p, q) = \frac{\max\limits_{x \in T_p \cup T_q}\mathrm{IC}(x)}{\mathrm{IC}_\mathrm{max}}
$$


<pre class="r">
group_sim(dag, group1, group2, method = "GroupSim_SimCHO")
</pre>


## GroupSim_SimALD

The similarity is calculated as:

$$ \mathrm{GroupSim}(p, q) = \max\left\{ 1 - \frac{\sigma(x)}{C_\mathrm{max}}: x \in T_p \cap T_q \right\} $$

<pre class="r">
group_sim(dag, group1, group2, method = "GroupSim_SimALD")
</pre>

# Set-based methods

Since $T_p$ and $T_q$ are two sets, the Kappa coeffcient, Jaccard coeffcient, Dice coeffcient and overlap coeffcient
can be naturally used.

<pre class="r">
group_sim(dag, group1, group2, method = "GroupSim_Jaccard", 
    control = list(universe = ...))
group_sim(dag, group1, group2, method = "GroupSim_Dice", 
    control = list(universe = ...))
group_sim(dag, group1, group2, method = "GroupSim_Overlap", 
    control = list(universe = ...))
group_sim(dag, group1, group2, method = "GroupSim_Kappa", 
    control = list(universe = ...))
</pre>


# Session info

```{r}
sessionInfo()
```


<script src="jquery.min.js"></script>
<script src="jquery.sticky.js"></script>
<script>
$(document).ready(function(){
    $("#TOC").sticky({
        topSpacing: 0,
        zIndex:1000    
    })
    $("#TOC").on("sticky-start", function() {

        $("<p style='font-size:1.2em; padding-left:4px;'><a id='TOC-click'>Table of Content</a></p>").insertBefore($("#TOC ul:first-child"));
        $("#TOC-click").hover(function() {
            $(this).css("color", "#0033dd").css("cursor", "pointer");
            $("#TOC").children().first().next().show();
            $("#TOC").hover(function() {
                $(this).children().first().next().show();
            }, function() {
                $(this).children().first().next().hide();
                $("body").off("hover", "#TOC");
            })
        }, function() {
            $(this).css("color", "#0033dd");
        })
        $("#TOC").children().first().next().hide();

    })
    $("#TOC").on("sticky-end", function() {
        $("#TOC").children().first().remove();
        $("#TOC").children().first().show();
    })
});
</script>

